\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-26 paper.tex
edits from JEre
tweaks

It is important that we make the distinction here between the local
trees that we can derive from a known ARG (as just discussed),
and the ARGs that we can derive from a sequence of \emph{estimated}
local trees.
The ARG inference methods \arbores~\citep{heine2018bridging} and
\espalier~\citep{rasmussen2022espalier} are illustrative in this context.
Both begin by splitting an input sequence alignment into
segments that are assumed to be non-recombining. Within
each segment, an initial local tree is estimated using
standard phylogenetic methods. By necessity, these local
trees will contain internal nodes that are unlabelled and
consist only of branching points: there is no information
shared between the independent tree estimation steps
across segments. Part of the task of stitching
these trees together
into an ARG is then, essentially, to generate labels for
the internal nodes, and decide which nodes persist
across multiple local trees. \arbores~bridges the gaps
between the inferred local trees via a coalescent-based
sampling scheme.
\espalier\ is specifically interested
in the details of recombination events,
and also attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination between these partially reconciled local trees.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.
% I don't really know if this is true, it's just a way of wrapping the
% section up by saying that other things we thought were settled might
% be worth thinking about again.
The combinatorial formulation of ARGs and local trees provided here
may help clarify these fundamental questions.

>>>>

It is important that we make the distinction here between the local
trees that we can derive from a known ARG (as just discussed),
and the ARGs that we can derive from a sequence of \emph{estimated}
local trees.
The ARG inference method
\espalier~\citep{rasmussen2022espalier} is illustrative in this context.
It begins by splitting an input sequence alignment into
segments that are assumed to be non-recombining. Within
each segment, an initial local tree is estimated using
standard phylogenetic methods. By necessity, these local
trees will contain internal nodes that are unlabelled and
consist only of branching points: there is no information
shared between the independent tree estimation steps
across segments. Part of the task of stitching
these trees together
into an ARG is then, essentially, to generate labels for
the internal nodes, and decide which nodes persist
across multiple local trees. 
% TODO add a sentence here saying how it does this. 
% that's what we're interested in illustrating here.
\espalier\ is specifically interested
in the details of recombination events,
and also attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination between these partially reconciled local trees.
A similar problem is tackled by \arbores~\citep{heine2018bridging}, which explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing and resimulating candidate SPR moves connecting two fixed, leaf-labelled local trees.
Exploration is done by varying the positions of these boundary trees along the sequence.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.
% I don't really know if this is true, it's just a way of wrapping the
% section up by saying that other things we thought were settled might
% be worth thinking about again.
The combinatorial formulation of ARGs and local trees provided here
may help clarify these fundamental questions.
